<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{fastinR: Fatty Acids and Stable Isotopes in R}
-->

Estimating diet proportions form Fatty Acids and Stable Isotopes
========================================================

DISCLAIMER: This is an evolving package and vignette, please read instructions [here](https://github.com/Philipp-Neubauer/fastinR/blob/master/README.md) for installation and to see what dependencies are required. If you find a bug or want so suggest improvements, please [submit an issue on github](https://github.com/Philipp-Neubauer/fastinR/issues?state=open), collaborations and contributions are very welcome!

DISCLAIMER 2: This tutorial is by no means a real analysis, which should proceed more carefully, and with longer MCMC runs. The example is designed to run through in a few minutes, and is therefore only an illustration of how an analysis of real data could proceed.

```{r echo=FALSE}
library(knitr)
options(markdown.HTML.header = 'vignette.css')
options(warn=-1)
```
```{r}
library(fastinR,warn.conflicts=F)
```
We start with a very simple simulated example. The easiest way to simulate relevant data is through the built in simulation GUI, which is called from the command line once fastinR has been loaded. Plese type ?simulation to obtain help on simulating your own 'dataset'. 

The package comes with a simulated dataset which can be called with `r data('Sim')`, which loads an object called _datas_ into the current environment.

Loading data from files
-----------------------

Raw data should be stored in .csv files in a prescribed format, see `?add_FA`, `?add_SI` and `?add_Covs` for details on formatting. You can also inspect saved output from `simulation()` to get a better idea of the correct file format.

Files are imported with the help of three constructor functions: `add_FA` to import fatty acid data, `add_SI` to import Stable Isotope data and `add_Covs` to import data on covariates or groupings that may be influencing predator diets. 

The package comes with simulated data in .csv files, which can be found using the `system.file` function. The `add_SI` constructor takes separate files for predator and prey Stable Isotope data, as well as files for additive fractionation coefficient means and variances - these are optional and can be specified manually in the function call, see the function help for details.

```{r}
SI.predators <- system.file("extdata", "Simdata_SI_preds.csv", package="fastinR")
SI.preys <- system.file("extdata", "Simdata_SI_preys.csv", package="fastinR")
Frac.Coeffs.mean <- system.file("extdata", "Simdata_SI_fc_means.csv", package="fastinR")
Frac.Coeffs.var <- system.file("extdata", "Simdata_SI_fc_var.csv", package="fastinR")

dats <- add_SI(SI.predators,SI.preys,Frac.Coeffs.mean,Frac.Coeffs.var)
```

We can now visualize the data on an Non-Metric Dimensional Scaling plot like so:

```{r fig.width=7, fig.height=6}
dataplot(dats)
```

The dats object now has a set of data in the right format for further analysis. We'll add fatty acids before we proceed. As with `add_SI`, the `add_FA` constructor takes separate files for predator and prey fatty acid data, conversion coefficient means and variances as well as fat content - these are again optional and can be specified manually in the function call, see the function help for details.

```{r}
FA.predators <- system.file("extdata", "Simdata_FA_preds.csv", package="fastinR")
FA.preys <- system.file("extdata", "Simdata_FA_preys.csv", package="fastinR")
Conv.Coeffs.mean <- system.file("extdata", "Simdata_FA_cc_means.csv", package="fastinR")
Conv.Coeffs.var <- system.file("extdata", "Simdata_FA_cc_var.csv", package="fastinR")
fat.conts <- system.file("extdata", "Simdata_fat_cont.csv", package="fastinR")

dats <- add_FA(FA.predators,FA.preys,fat.conts,Conv.Coeffs.mean,Conv.Coeffs.var,datas=dats)
```


*Note* the last argument in `add_FA` now contains a reference to the data object that was constructed from the Stable Isotopes beforehand. `add_FA` just adds fatty acid data to the mix - the same would apply if data were added the other way around (only `add_Covs` works separately, it needs it's own object).

Plotting the joint dataset:

```{r fig.width=7, fig.height=6}
dataplot(dats)
```

Data Grooming
-------------

The grooming step in this case corresponds to selecting relevant variables for the analysis. In practice, one could measure lots of Fatty Acids, and researchers often choose arbitrary cutoff points in proportions to reduce the dataset. However, even Fatty Acids occurring in low proportions may be useful for source discrimination and diet estimation if they introduce systematic differences between sources (prey items). 
The `select_vars` function provides a graphical way to choose a subset of Fatty Acids according to their contribution to prey separation and reduction in collinearity in the prey matrix. Using `select_vars` in the simulated example shows that only really 2 Fatty Acids contribute to source separation - the plot of cumulative separation converges to one after adding the two most important fatty acid proportions (see also cumulative separation scores displayed in the console): all remaining FAs only add to collinearity in the source matrix. We choose F Atty Acid no 3,2 and 6 for analysis (the second call does this directly without the graphical interface).

```{r,eval=FALSE}
dats.subset <- select_vars(dats,plot=F)
```
or
```{r }
dats.subset <- select_vars(dats,c(3,2,6))

# inspecting
dats.subset$datas.FA$n.fats
```

The data object `dats` is now ready for analysis, we can plot the dataset with it's new subset of fatty acids as before:

```{r fig.width=7, fig.height=6}
dataplot(dats.subset)
```

Apart from a rotation, the overall configuration should be similar.

Bayesian Analysis
-----------------

The actual analysis usues Markov Chain Monte Carlo to estimate posterior distributions for diet parameters. Estiamtion can be performed locally (in the active R session) or in a distributed way, using as many R sessions as Markov Chains. A good strategy is usually to run a few chains locally for short runs, and then run 3 or more  chains in parallel in a distributed way once a satisfactory set of priors, thinning interval etc has been found (setting the number of chains to a maximum of n-1 cores on your CPU is a good idea to not hog ALL resources). The `run_MCMC` functions sets up the MCMC runs and takes `spawn = T` or `spawn = F` as parameter to run chains in R slave processes or locally, respectively.

The Rnot parameter sets the prior scale for the predator logistic normal covariance matrix, and will often determine how well the chain mixes, that is, how well it explores the parameter space, or if it gets stuck in local modes (that can be of low probability). In the latter case, one would see strong autocorrelation in the Markov Chains for individual paramters. Increasing ```Rnot```/```Rnot_SI``` can help, but will make it ahrder to get precise estimates. There is thus a tradeoff between accuracy and mixing here, and often one will just have to run an analysis for many iterations and with a large thinning interval (e.g., >100k iterations with a thinning interval of >100). This should be done after finding reasonable parameters for Rnot on a shorter run, and then letting the final analysis run with long chains.

For a combined analysis of stable isotopes and fatty acids, it is often useful to run the two datatypes separately to assure that good priors can be found for both datasets independently, and then combining them for a final analysis. The analysis type is chosen in the appropriate option in the function call.

### Estimating population proportions

#### From Stable Isotopes alone

Lets start with an analysis of the stable isotopes, estimating only global (population) level diets. We will use the default prior on the predator covariance matrix, and will adjust this prior subsequently. *WARNING* This might take a while depending on your resources, the size of the dataset and the parameters used for the MCMC. If you are not familiar with Markov Chain Monte Carlo, please take a minute to read up on it, there are lots of great explanations and summaries on the web.

```{r}
Pop.SI <- run_MCMC(datas=dats.subset,nIter=10000,nBurnin=1000,nChains=3,nThin=10,Data.Type='Stable.Isotopes',Analysis.Type='Population.proportions',Rnot_SI=0.1,plott=F,spawn=F)
```

Plotting the MCMC is the easiest way to ensure that the sampler is mixing - meaning that the chain explores the posterior distribution of each parameter efficiently.

```{r fig.width=7, fig.height=6}
MCMCplot(Pop.SI)
```

The mixing isn't great, meaning that some autocorrelation is evident in the Markov Chain time series in left-hand plots. The `diags` function gives more information, displaying two types of diagnostics in the console.:

```{r}
diags(Pop.SI)
```

Based on the diagnostics, it seems that we're doing OK, but that the Stable Isotopes don't give much certainty about diet proportions:

```{r}
summary(Pop.SI)
```
```{r fig.width=7, fig.height=6}
plot(Pop.SI,save=F)

```

The credible intervals are very large for Prey items 1 and 2, and the correlation plot suggests that there is a reasonably strong posterior correlation between estimates of prey items 1 & 2.

Increasing the default prior ```Rnot_SI``` on the predator covariance matrix sometimes leads to better mixing. We'll also run more iterations and set `spawn=T` for this.

```{r}
Pop.SI2 <- run_MCMC(datas=dats.subset,nIter=100000,nBurnin=10000,nChains=3,nThin=100,Data.Type='Stable.Isotopes',Analysis.Type='Population.proportions',Rnot_SI=1,plott=F,spawn=T)
```

```{r fig.width=7, fig.height=6}
MCMCplot(Pop.SI2)
```

```{r}
diags(Pop.SI2)
```

Mixing doesn't seem to get much better, indicating that this is probably as good as it gets with stable isotopes alone.

#### Fatty Acids alone

Lets repeat this with Fatty Acids, again starting with the default prior. Note that `spawn=T` this time to save some time.

```{r}
Pop.FA <- run_MCMC(datas=dats.subset,nIter=10000,nBurnin=1000,nChains=3,nThin=10,Data.Type='Fatty.Acid.Profiles',Analysis.Type='Population.proportions',Rnot = 0.2,plott=F,spawn=T)
```

```{r fig.width=7, fig.height=6}
MCMCplot(Pop.FA)
```

Once again the mixing isn't great:

```{r}
diags(Pop.FA)
```

Again not too bad according to the diagnostics, although the chains should be run for longer (see Raftery-Lewis diagnostics ) - but lets try again with `Rnot = 1`

```{r}
Pop.FA2 <- run_MCMC(datas=dats.subset,nIter=10000,nBurnin=1000,nChains=3,nThin=10,Data.Type='Fatty.Acid.Profiles',Analysis.Type='Population.proportions',Rnot = 1,plott=F,spawn=T)
```

```{r fig.width=7, fig.height=6}
MCMCplot(Pop.FA2)
```

```{r}
diags(Pop.FA2)
```

Looks very similar, so lets do a final run with 30k iterations as suggested by the `diags` output and a thinning interval of 30 should be better. (Note that values suggested by the diagnostics may vary from one run to the next. Also note that we already ran 30k as we ran 3 parallel chains here...).

```{r}
Pop.FA3 <- run_MCMC(datas=dats.subset,nIter=30000,nBurnin=1000,nChains=3,nThin=30,Data.Type='Fatty.Acid.Profiles',Analysis.Type='Population.proportions',Rnot = 1,plott=F,spawn=T)
```

```{r fig.width=7, fig.height=6}
MCMCplot(Pop.FA3)
```

```{r}
diags(Pop.FA3)
summary(Pop.FA3)
```
```{r fig.width=7, fig.height=6}
plot(Pop.FA3,save=F)
```
There is still substantial uncertainty about the diet proportions: it looks as though Prey 1 is a dominant source, but the credible intervals are large. Let's try combining stable isotopes and fatty acids:

#### Combining fatty acids and stable isotopes

```{r}
Pop.Combined <- run_MCMC(datas=dats.subset,nIter=30000,nBurnin=1000,nChains=3,nThin=30,Data.Type='Combined.Analysis',Analysis.Type='Population.proportions',Rnot = 1,Rnot_SI=1,plott=F,spawn=T)
```

```{r fig.width=7, fig.height=6}
MCMCplot(Pop.Combined)
```
```{r}
diags(Pop.Combined)
summary(Pop.Combined)
```
```{r fig.width=7, fig.height=6}
plot(Pop.Combined,save=F)
```

The combined analysis reduces uncertainty slightly, especially for source 3, but sources remain uncertain with somewhat large credible intervals. 

To compare the three approaches explicitly, we can use `multiplot`. For `multiplot` the three results need to be combined in a list:

```{r}
Pop.list <- list(Stable.Isotopes = Pop.SI2, Fatty.Acids = Pop.FA3, Combined = Pop.Combined)

multiplot(Pop.list,save=F)
```

The original data was simulated to include groups of predators with rather different proportions (visible in the dataplots above). Lets have a look at the proportions that were used to simulate the dataset:

```{r}
proportions.path <- system.file("extdata", "Simdata_props.csv", package="fastinR")
proportions <- read.csv(proportions.path,header=F,row.names=1)

colnames(proportions) <- unique(dats$prey.ix)

proportions

# overall mean proportions
colMeans(proportions)
```

Looking at the proportions, it is clear that there are two groups of predators (e.g., juveniles and adults), one group preys preferentially on Prey 1 while the other preys mostly on Prey 2. Let's see if we can pick this up by estimating individual predator proportions:

### Estimating individual proportions

Lets try this with fatty acids first - given that there was limited information about diets in the stable isotopes alone, they can't be expected SI alone to provide the extra information we're after. 

We now need to deal with one extra prior to be set: the prior for the variance of diet proportions. After some exploratory runs, `even=2` seems like a reasonable compromise between our ability to detect differences (facilitated for smaller values of `even`) and the ability of the Markov Chains to mix properly (easier for larger `even` - see the documentation of `run_MCMC`).

#### From Fatty Acids

```{r}
Ind.FA <- run_MCMC(datas=dats.subset,nIter=10000,nBurnin=1000,nChains=3,nThin=10,Data.Type='Fatty.Acid.Profiles',Analysis.Type='Individual.proportions',Rnot=1,even=2,plott=F,spawn=T)
```

We won't show the output of the next commands anymore since it is far too long. The commands are the same as for a population proportion analysis:

```
MCMCplot(Ind.FA)

diags(Ind.FA)

summary(Ind.FA)
```
```{r}
plot(Ind.FA,save=F)
```

Despite this MCMC run being an order of magnitude too short, we can see that the grouped pattern emerges quite clearly in the last plot. Nevertheless, posterior correlations remain for prey items 1&2 remain strong (we'll soon see why!). First, let's see if the grouping becomes more pronounced with the inclusion of stable isotopes:

#### Combined analysis

```{r}
Ind.Combined <- run_MCMC(datas=dats.subset,nIter=10000,nBurnin=1000,nChains=3,nThin=10,Data.Type='Combined.Analysis',Analysis.Type='Individual.proportions',Rnot=1,Rnot_SI=1,even=2,plott=F,spawn=T)
```
Again skipping diagnostics output:

```
MCMCplot(Ind.Combined)

diags(Ind.Combined)

summary(Ind.Combined)
```
```{r}
plot(Ind.Combined,save=F)
```

The uncertainty for predators preying predominantly on Prey 2 is significantly reduced (but once again the analysis should be run for an order of magnitude longer with a larger thinning interval to ensure that the estimates are reliable).

While the patterns here offer great insights into individual diet proportions, they do not provide a means to estimate the population distribution of diet proportions for these two groups. This can be achieved in an anova type linear model:

### Estimating group proportions

We now need to add covariates in the form of group membership. The reasoning and procedure is the same for continuous covariates (e.g., size). To add the covariates, we use the `add_Covs` constructor:

```{r}
group.path <- system.file("extdata", "Simdata_groups.csv", package="fastinR")
Covs <- add_Covs(Groups=group.path)


Cov.Combined <- run_MCMC(datas=dats.subset,Covs=Covs,nIter=100000,nBurnin=10000,nChains=3,nThin=100,Data.Type='Combined.Analysis',Analysis.Type='Analysis.with.Covariates',Rnot=1,Rnot_SI=1,even=2,plott=F,spawn=T)
```

```
MCMCplot(Cov.Combined)

diags(Cov.Combined)

summary(Cov.Combined)
```
```{r}
plot(Cov.Combined,save=F)
```

We can now clearly see the difference between the two group's diet proportions!