Analysis of Stowasser et al. 2006 data
========================================================

Here we demonstrate the validity of our methods by re-analysing the experimental dataset of Stowasser et al. The original paper investigated the evolution of stable isotope (C/N) and fatty acid signatures for squid fed either fish, crustacean or mixed diets, as well as switched diet regimes. Our specific aim here is to estiamte diet proportions of mixed diet individuals, and to show the advantage of using both markers in concert over using a single marker for diet analysis. 

We begin the analysis using just SI. Hussey et al 2014 (Ecology Letters Volume 17, Issue 2, pages 239–250, February 2014), following earlier anlyses by Caut et al and others, showed that discimintation is prey N dependent, and used a meta-analytic model to estiamte discrimination coefficients. 

We initially estiamte prey specific discimination in stable isotopes following Hussey et al 2014, and develop analogous methods for fatty acid signatures. Discrimination is estiamted for both types of marker from animals that were fed on diets containing a single prey type (Fish or Crustaceans).

We first read in the data tables for SI:

```{r}
prey.table <- read.csv('prey_SI.csv',header=T,stringsAsFactors=F)
```
We combine fish species since they were not differentiated in treatments:

```{r}
prey.table[prey.table[,1] != 'Grass shrimp',1] <- 'Fish'
}

pred.table <- read.csv('predator_SI.csv',header=T,stringsAsFactors=F)
pred.ix <- pred.table[,1]
pred.table <- pred.table[,2:3]
```
We estimate discrimintation coefficients (also termed fractionation or conversion for SI and FA, respectively) using Bayesian models. The analysis is done in a separate file (Discrim.Analysis.R) and produces files discr.means.csv/discr.var.csv for SI, and corresponding cc_FA.csv and cc_FA_var.csv for FA.The estiamtion isn't trivial, so we just pass over it here. We instead proceed with the analysis by adding the necessary items for the final analysis to the workspace, using the add_SI function in fastin-R. This function just adds the data, and puts it into a specific format.

```{r}
squid.data <- add_SI(SI.predators=pred.table,SI.preys=prey.table,Frac.Coeffs.mean='discr.means.csv',Frac.Coeffs.var='discr.var.csv')
```

First, we need to calculate the moments for combined fat content of prey types, since they are given for individual species in the original paper. Since we do not have the original fat content data (could ask for it?) we simply combine them as a mixture distribution:

```{r}

# load fatty acid data tables 
prey.table.FA <- t(read.csv('prey_FA.csv',header=T,stringsAsFactors=F,row.names=1))
pred.table.FA <- t(read.csv('predator_FA.csv',header=T,stringsAsFactors=F,row.names=1))


Silverside <- 2.27
Silverside.sd <- 1.02 

Sailfin.Molly <- 3.97
Sailfin.Molly.sd <- 1.25

Sheepshead.Minnow <- 3.59 
Sheepshead.Minnow.sd <- 1.45

Striped.Mullet <- 1.29 
Striped.Mullet.sd <-  0.53

fish.means <- c(Silverside,Sailfin.Molly,Sheepshead.Minnow,Striped.Mullet)

fish.vars <- c(Silverside.sd,Sailfin.Molly.sd,Sheepshead.Minnow.sd,Striped.Mullet.sd)^2

#numbers for each species
n.Silverside <- length(grep('Silverside',rownames(prey.table.FA)))
n.Sailfin.Molly <- length(grep('Sailfin.Molly',rownames(prey.table.FA)))
n.Sheepshead.Minnow <- length(grep('Sheepshead.minnow',rownames(prey.table.FA)))
n.Striped.Mullet <- length(grep('Striped.Mullet',rownames(prey.table.FA)))

#into vector
n.fish <- c(n.Silverside,n.Sailfin.Molly,n.Sheepshead.Minnow,n.Striped.Mullet)

#sum
total.fish = n.Silverside + n.Sailfin.Molly + n.Sheepshead.Minnow + n.Striped.Mullet
```

Assuming a log normal model for fat content - this is done internally in fastin-R, so the next bit is for illustrative purposes only...

```{r}
fish.ln.vars = log(fish.vars + fish.means^2) - 2*log(fish.means)
fish.ln.means = log(fish.means)-fish.ln.vars/2

hist(rlnorm(10000,meanlog=fish.ln.means[1],sdlog=sqrt(fish.ln.vars[1])),30)

#grouping fish species using weighted mean of fish
fish.ln.mean <- sum((n.fish/total.fish)*fish.ln.means)

## Var = \Sum w_i((\mu_i-mu)^2+SD_i^2)

fish.ln.var <- sum((n.fish/total.fish)*((fish.ln.means-fish.ln.mean)^2+fish.ln.vars))

hist(rlnorm(10000,meanlog=fish.ln.mean,sdlog=sqrt(fish.ln.var)),30)

shrimp.mean <- 2.43
shrimp.var <- 0.66^2

shrimp.ln.var = log(shrimp.var + shrimp.mean^2) - 2*log(shrimp.mean)
shrimp.ln.mean = log(shrimp.mean)-shrimp.ln.var/2

hist(rlnorm(10000,meanlog=shrimp.ln.mean,sdlog=sqrt(shrimp.ln.var)),30)

fat.cont <- rbind(c(fish.ln.mean,fish.ln.var),c(shrimp.ln.mean,shrimp.ln.var))

write.csv(fat.cont,file='fat.cont.csv')

```
We estimate discrimintation coefficients (also termed fractionation or conversion for SI and FA, respectively) using Bayesian models. The analysis is done in a separate file (Discrim.Analysis.R) and produces files discr.means.csv/discr.var.csv for SI, and corresponding cc_FA.csv and cc_FA_var.csv for FA.The estiamtion isn't trivial for FA, so we just pass over it here. We instead proceed with the analysis by adding the necessary items for the final analysis to the workspace.

```{r}
add_FA(FA.predators=pred.table.FA,)


```

You can also embed plots, for example:

```{r fig.width=7, fig.height=6}
plot(cars)
```

